Stagnating State Funding for Higher Education and its Effect on Faculty at US Universities

### 2023-07-13 text/sections/01-introduction.tex
re-writing

Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) over the last three decades.
At the same time, both public university student enrolment and employment of contingent lecturers dramatically increased at public universities.
I show that four-year, degree-granting public universities have experienced a systematic fall in state funding per student, and this fall led to a substitution away from tenure-track/tenured professors towards contingent lecturers.
Falls in state funding for public universities, instrumented by state-level shocks to state-wide higher education, led to a fall in the number of tenure-track/tenured professors per student within a university, and an increase in the number of lecturers.
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors, so that the stagnating state support for public universities has implications for the wider structure of US higher education and research.

@@

I employ a shift-share instrument to identify changes in state funding for higher education, which exploits how universities differ in their financial reliance on state funding, combined with yearly shocks to the amount an entire state funds higher education --- following \cite{NBERw23736,NBERw27885}.
State governments decide how to fund higher education, and everything else, by a complicated process which can be influenced by local economic conditions, changing state priorities, or even lobbying from the state universities themselves.
As such, it is necessary to employ an identification strategy, such as the shift-share instrument, to account for endogeneity in universities' state funding.
The shift-share instrument interacts reliance on state funding in a base period with the entire state's total funding per higher education student for each year, to exploit changes in funding and how much each university historically relied on state funding for higher education.

\textbf{TODO: Write section on the data sources.  See Lovenheimn (2020) on how to do it.}

Falls in state funding affects faculty composition at public universities, which substitute away from tenured and tenure-track professors towards contingent lecturers.
A fall of $-10$\% in state funding leads to a fall of $-1.4$\% in the number of assistant professors per student at a university, a fall of $-1.2$\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock, showing that the effect is not isolated to the immediate year of the first funding shock. 
Over the same time period, state funding fell by around $-$35\%, while the count of professors per student fell by $-$9\%, so that these results show that falls in state funding explain around a third of the substitution away from tenure-track and tenure professors towards contingent lecturers.
Additionally, I use individual level data on all Illinois public university professors in 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by the falls in state funding (and by the funding shocks).
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or leaving) professors.

A number of studies have shown that stagnating state funding for higher education negatively impacts student outcomes.
\cite{NBERw23736} show that increases in public university spending (via funding shocks) increases enrolment and degree completion among students,\footnote{
    \cite{miller2022making} further analyse the effects of falls in university revenues, by finding that reductions in public universities tuition prices (suggestively motivated by falling state funding) leads to reductions in provided financial aid.
}
that high school students exposed to similar public school funding shocks are less likely to attend college \citep{jackson2021school}, and that these shocks induce public universities to rely more on tuition as a source of revenue \citep{bound2019public}.
\textbf{Faculty outcomes surveyed by \cite{ehrenberg2003studying}.}
\cite{NBERw27885} use the same methodology to show that increases in state funding lead to degree lower completion time and later-life debt for students.\footnote{
    \cite{bound2007cohort} use similar variation in state funding per student to show that lower funding leads to lower completion rates at public universities.
}

@@

While it is not immediately clear whether a systematic substitution towards lecturers away from professors reduces teaching effectiveness, it is clear that stagnating state funding has negatively impacted student outcomes \citep{NBERw23736,NBERw27885}.
\textbf{REWRITE:} where faculty composition is a possible mechanism for how the stagnating state funding for higher education transmits to students, and has led to worsening education outcomes at public universities for the last thirty years.

\textbf{TO-DO:
    This paper precedes as follows.
}

>>>>

Public universities educate the majority of higher education students in the US, yet have experienced a stagnation in state funding over the last three decades .
At the same time, public university employment of lecturers dramatically increased at public universities, while enrolment rose dramatically.
Previous research has examined the effects of the decline in state funding on student outcomes \citep{NBERw23736,NBERw27885}, but there is little evidence of the effects of thid funding decline on faculty.
I show that four-year, degree-granting public universities have experienced a systematic decline in state funding per student, and use a shift-share instrument to show that this fall led to a substitution away from tenure-track/tenured professors towards contingent lecturers.
Analsysis of every public university faculty member in the state of Illinois shows that incumbent faculty were relatively unaffected, implying that changes in faculty composition arose by disrupting public universities' ability to hire or replace their tenure-track/tenured faculty. 
Private universities were not exposed to similar financial constraints during this same time period, and do not exhibit the substitution away from tenure-track/tenured professors, so that the stagnating state support for public universities has implications for the wider structure of US higher education and research.

There are various margins through which state fundings cuts can impact a university's faculty.
A university's level of resources affect the ability that departments can hire new professors, retain their current fauclty, or provide retirement incentives.
So that it is natural to expect that funding cuts may affect the composition of professors at a university.
Similarly, the wages a university pays their faculty is affected by funding for the university, including the starting salary it offers to new professors and the amount of yearly raises (if any) for incumbent faculty.
So that it is natural to expect that systematic funding cuts will lead to changes in faculty composition, and possibly affect incumbent professors, too.
Yet, it is a priori unclear whether/how much each of these margins adjusts to funding cuts, as these decision depend on how the university administrations view the productivity of different inputs, plus the degree of wage and employment stickiness over time.

@@

I use data from the Integrated Postsecondary Education Data System (IPEDS) to study the impact of changes in state funding for US public univeristies 1990--2017.
I combine these data with panel data for every public university faculty member in the state of Illinois for 2010--2021 (Illinois Board of Hdigher Education Database, IBHED), to measure how faculty are affected by changes in state funding.
These data allow me to answer the following research question: if a US public university receives an extra \$1,000 for each of their students (or a 10\% increase), then do they adjust the number of and composition of faculty they employ, and how does this affect individual faculty already at the university?

I employ a shift-share instrument to identify changes in state funding for higher education, which exploits how universities differ in their financial reliance on state funding, combined with yearly shocks to the amount an entire state funds higher education --- following \cite{NBERw23736,NBERw27885}.
State governments decide how to fund higher education, and everything else, by a complicated process which can be influenced by local economic conditions, changing state priorities, or even lobbying from the state universities themselves.
As such, state funding is not randomly assigned to universities and so it is necessary to employ an identification strategy, such as the shift-share instrument, to account for this.
The shift-share instrument interacts reliance on state funding in a base period with the entire state's total funding per higher education student for each year, to exploit changes in funding and how much each university historically relied on state funding for higher education.

\textbf{Adjust this to have dollar figures, and generall re-do desc of results.}
Falls in state funding affects faculty composition at public universities, which substitute away from tenured and tenure-track professors towards contingent lecturers.
A fall of $-10$\% in state funding leads to a fall of $-1.4$\% in the number of assistant professors per student at a university, a fall of $-1.2$\% for tenured professors, and an increase in 4.4\% the number of lecturers per student.
Local projection estimates show that these effects linger for three to four years after the initial funding shock, showing that the effect is not isolated to the immediate year of the first funding shock. 
Over the same time period, state funding fell by around $-$35\%, while the count of professors per student fell by $-$9\%, so that these results show that falls in state funding explain around a third of the substitution away from tenure-track and tenure professors towards contingent lecturers.
Additionally, I use individual level data on all Illinois public university professors in 2010-2021 to investigate whether the shocks to state funding affected individual professors.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system.
Yet the hiring rate for new professors at public universities was negatively impacted by the falls in state funding (and by the funding shocks).
This implies that faculty composition change arose by limiting the hiring of new professors; public universities increasingly hired contingent lecturers, and increasingly did not replace their retiring (or leaving) professors.

\textit{Re-do the lit review to focus on mechanisms of student quality.}
A number of studies have shown that stagnating state funding for higher education negatively impacts student outcomes.
\cite{NBERw23736} show that increases in public university spending (via funding shocks) increases enrolment and degree completion among students,\footnote{
    \cite{miller2022making} further analyse the effects of falls in university revenues, by finding that reductions in public universities tuition prices (suggestively motivated by falling state funding) leads to reductions in provided financial aid.
}
that high school students exposed to similar public school funding shocks are less likely to attend college \citep{jackson2021school}, and that these shocks induce public universities to rely more on tuition as a source of revenue \citep{bound2019public}.
\textbf{Faculty outcomes surveyed by \cite{ehrenberg2003studying}.}
\cite{NBERw27885} use the same methodology to show that increases in state funding lead to degree lower completion time and later-life debt for students.\footnote{
    \cite{bound2007cohort} use similar variation in state funding per student to show that lower funding leads to lower completion rates at public universities.
}

@@

Mine is the frst paper to provide causal evidence of the impact of state funding for higher education on faculty composition and wages.
The closest related research has examined (\textbf{theoretical papers here}) $\hdots$

My results provide suggestive evidence on the mechanisms for previous research on the negative effects of state funding cuts on student outcomes at US public universities \citep{zhu2021limited,ehrenberg2012american}.

Lastly, my results providde evidence that the long term trends in higher education funding and of non-tenure track faculty are causally related.
\textbf{Sentence on how they will continue.}

This paper proceeds as follows.
\autoref{sec:data} describes the data for university finances and faculty in Illinois, and trends in public university funding for the last three decades.
\autoref{sec:conceptual} gives the conceptual framework for how state funding may affect faculty.
\autoref{sec:empirics} draws the empirical framework for isolating the causal effects of state funding on faculty composition and individual faculty, and \autoref{sec:results} presents the empirical results.
\autoref{sec:discussion} discusses the context and implications for the findings.
\autoref{sec:conclusion} concludes.

### 2023-07-13 text/sections/04-empirical.tex
re-writing

The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} presents results of the first-stage regression, separately with and without a control for tuition revenue per student, plus institution and year fixed effects;
Panel A shows estimates of a funding shock of \$1 per student in the state on state funding per student at the university, and Panel B the \% increase effects of a 1\% increase in the shock per student on state funding per student.
\textbf{TO-DO:
    Add description the dollar-by-dollar, of the first stage.
}
We note columns (1) and (2) estimate that a funding shock (per student in the entire state) of 10\% is associated with 9.8\% change in state funding per student at the university; the instrument is strong, and we note similarity in estimates with and without inclusion of fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider or bad control.
Column (3) shows similar estimates to columns (1), (2) thanks to inclusion of fixed effects, so that the fixed effects was effective in soaking variation in per-student tuition revenues at the institution-year level.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with.

>>>>

The shift-share instrument has a strong first stage for universities' yearly state funding.
\autoref{tab:firststage-reg} presents results of the first-stage regression, separately with and without a control for tuition revenue per student, plus institution and year fixed effects;
Panel A shows estimates of a funding shock of \$1 per student in the state on state funding per student at the university, and Panel B the \% increase effects of a 1\% increase in the shock per student on state funding per student.
Columns (1) shows that a funding shock of \$10 (per student in the entire state) is associated with \$11.76 less state funding per student at the univeristy, with the corresponding -10\% funding shock leads to -9.77\%
less funding (coumn 1, panel B), with similar estimates with and without including fixed effects.
Column (2) shows estimates of the first-stage without including fixed effects, and gives less precise estimates for the funding shock, likely thanks to systematic differences in universities unaccounted for without fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider (or bad control).
Column (3) shows similar estimates to column (1) in both Panels A and B thanks to inclusion of fixed effects, so that the uncertainty in including tuition revenue as a possible bad control for the level of state funding does not matter thanks to the inclusion of fixed effects.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with.
